\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-23 paper.tex
Some edits from Gregor

\section{Genome ARGs}
\label{sec-gARG}
We define a genome as the complete set of genetic material that a child
inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
We will also use the term ``genome'' in its
more common sense of ``the genome'' of a species, denoting the full set
of chromosomes, and hope that the distinction
will be clear from the context.
We are not concerned here with mutational processes or observed sequences,
but consider only processes of inheritance,
following the standard practice in coalescent theory.
We also do not consider structural variation, and assume that all
samples and ancestors share the same genome coordinate space.

A genome ARG (gARG) is a directed acyclic graph in which nodes represent
haploid genomes and edges represent
genetic inheritance between an ancestor and a descendant.
The topology of a gARG specifies that genetic inheritance
occurred between particular
ancestors and descendants, but the graph connectivity
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination
(and other processes that lead to the partial genome inheritance)
 we ``annotate'' the edges with the genome
coordinates over which inheritance occurs.
We can define a gARG formally as follows.
Let $N \subset \mathbb{N}$ be the set of nodes representing
the genomes in the gARG, $S \subseteq N$ be the set of sampled genomes.
% (but see Section~\ref{sec-retro-pro} for some subtlety around this point).
Then, $E$ is the set of edges, where each element
is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals
over which genome $c$ inherits from $p$.
Thus, each topological connection between
a parent and child node in the graph is annotated with a set of
inheritance intervals $I$.
Here, the terms parent and child are used in graph sense;
these nodes respectively represent ancestor and descendant genomes,
which can be separated by multiple generations.
We will use these two sets of terms interchangeably.

In many settings, nodes are dated, i.e.\ each
node $u\in N$ is associated with a time $\tau_u$.
The topological relationships
defined by the directed graph structure, however,
 is sufficient for many applications,
and we will assume that node dates are not known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata such as sample identifiers or quality-control metrics
with nodes; it is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.Concretely defining nodes as integers in this way has some
significant advantages. In particular, it provides a direct link
between an ARG and its local trees in terms of a well-defined
combinatorial object (Section~\ref{sec-ARG-and-local-trees}),
as well as enabling straightforward connection between
definitions and practical implementation (Section~\ref{sec-efficiency}).

@@

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The pedigree relationships between the diploid individuals is
shown, along with their paths
of genetic inheritance and constituent genomes (labelled
with lowercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic
intervals).
Thus, individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the genomes
% FIXME "contained within" is a poor phrase here, use "constituent" or something
contained within a parent.
In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant product of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations (A), the corresponding gARG (B), and finally the corresponding
sequence of local trees  along the
genome (C).
% Jerome: this feels unecessary and "genealogy" is often interpreted as
% "pedigree" so needs to be used very carefully.
% The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
% which changes between local trees due to recombination.
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes.
See Section~\ref{sec-ARG-and-local-trees} for further details on
how the local trees are embedded in an ARG.
% Following the above definitions, corresponding data encoding for
% this gARG is shown in Tab.~\ref{tab-gARG-data}.

@@

\section{Event ARGs}
\label{sec-eARG}
% % I feel like sophisticated readers would at this point be saying,
% % ah, yeah, that's basically what an ARG is just in different words;
% % what's the point of this paper?
The genome ARG framework defined in the previous section is
in many ways a clarification of existing treatments
\cite[e.g.][]{mathieson2020ancestry,shipilina2023origin},
adding concrete details to the mechanisms by which
differential inheritance of genetic material are modelled.
% This is definitely needed for the Nicks, who are always thinking
% about the deeper processes and are quite happy being loosey-goosey
% with representation details
Note that here, and throughout, we are not questioning the form of the
actual ancestral processes that occur in nature, but rather how we
\emph{represent} such realised histories in a practical sense.
% Again, need to remind the reader that there's an actual point
% to all this getting stuck into the details.
These practical details,
as demonstrated through the remainder of the article, have important
consequences not only for how methods exchange information about
simulated and inferred ARGs, but more fundamentally in
how we set our goals for inference and evaluate the success of results.

@@

% The prevailing view is that ARGs are composed of events, and so we
% characterise this view
% Besides viewing ARGs as being composed of events, another
% feature of the definitions given above is that they do not
% explicitly describe
% % TODO check this! Do they talk about how trees are embedded?
% the details of how such events are represented.
Where the concrete details of an ARG data structure
(the encoding) are
% TODO more?
specified~\citep[e.g.][]{
rasmussen2014genome,gusfield2014recombinatorics,ignatieva2021kwarg}
they follow the approach described by Griffiths and colleagues.
Because it is focused on recording the details of events,
we will refer to this Griffiths encoding as an ``Event ARG'', or eARG.
In an eARG, we have two different types of node in the graph
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoint, and these breakpoints
are used to reconstruct the local trees.
This is done by tracing rootwards through the graph from the samples,
taking decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}.
Fig.~\ref{fig-event-arg} shows an example of a classical
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
We assume here that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

While the Griffiths approach of annotating recombination nodes
with a breakpoint in an eARG is a concise and elegant way of describing
realisations of the coalescent, it has some significant
issues when applied to the complexities of present-day datasets.
Because we are explicitly modelling the \emph{properties} of events,
the encoding must be extended to incorporate
different types of event that are not simple crossover recombination
or common ancestry.
Processes such as gene conversion, for example, can in principle
be included, but either require additional node types
or (e.g.) multiple recombination nodes joined by a zero-length edge.
Modelling multiple recombinant offspring of the same parents
(while keeping track of their identity as siblings)
as occurs
in Wright-Fisher models (e.g.\ Fig.~\ref{fig-arg-in-pedigree}),
or densely sampled multi-generational
data~\citep[e.g.][]{turnbull2018hundred,hayes20191000,RosFreixedes2022}
% circumlocutions isn't technically correct, but I like the roundabout
% overly verbose feel it gives.
would require more involved circumlocutions.
These and other problems can be worked around
(although, we would ideally like everyone to use the
\emph{same} workarounds if methods are to interoperate; see the Discussion),
but it is much simpler to use the gARG encoding described
in the previous section.
By recording the \emph{outcome} of events in terms of the
intervals of genetic inheritance on edges instead of recording the
properties of the events themselves, we remove the need to explicitly
model different types of event in the data structure.
More fundamentally, because gARGs
allow us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant,
we can encode the outcome of
\emph{multiple} genetic inheritance events, and to
systematically describe the uncertainty around their
ordering (see Section~\ref{sec-precision}).

@@

Any eARG can be converted to a sample-resolved gARG using the same machinery,
and the two-step process is illustrated in Fig.~\ref{fig-ancestry-resolution}.
The first step is to take the input eARG (Fig.~\ref{fig-ancestry-resolution}A)
and convert all event nodes to genome nodes and add inheritance annotations to
the gARG's edges (Fig.~\ref{fig-ancestry-resolution}B).
If the node is a common ancestor node, we annotate the single
outbound edge with the interval $[0,L)$, for a genome of length $L$. If the
node is a recombination event with a breakpoint $x$, we annotate the two
outbound edges respectively with the intervals $[0, x)$ and $[x, L)$. These
inheritance interval annotations are clearly in one-to-one correspondence with
the information in the input eARG. They are also also analogous to the
inheritance intervals we get on the edges in a prospective ARG
produced by a forwards-time simulation, which are concerned with recording
the local genetic relationship between a parent and child genome and not
necessarily minimal in terms of the resolved ancestral material of a sample.
Thus, the final step is to use the ``simplify'' algorithm to perform this
resolution~(Fig.\ref{fig-ancestry-resolution}C).

>>>>

\section{Genome ARGs}
\label{sec-gARG}
We define a genome as the complete set of genetic material that a child
inherits from one parent. A diploid individual  (we assume diploids here
for clarity, but the definitions apply to organisms of arbitrary ploidy)
% Q: will someone nitpick that we're assuming diploid obligate sexual species?
therefore carries two genomes, one inherited from each parent.
We will also use the term ``genome'' in its
more common sense of ``the genome'' of a species, denoting the full set
of chromosomes, and hope that the distinction
will be clear from the context.
We are not concerned here with mutational processes or observed sequences,
but consider only processes of inheritance,
following the standard practice in coalescent theory.
We also do not consider structural variation, and assume that all
samples and ancestors share the same genome coordinate space.

A genome ARG (gARG) is a directed acyclic graph in which nodes represent
haploid genomes and edges represent
genetic inheritance between an ancestor and a descendant.
The topology of a gARG specifies that genetic inheritance
occurred between particular
ancestors and descendants, but the graph connectivity
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination
(and other processes that lead to the partial genome inheritance)
 we ``annotate'' the edges with the genome
coordinates over which inheritance occured.
We can define a gARG formally as follows.
Let $N \subset \mathbb{N}$ be the set of nodes representing
the genomes in the gARG, $S \subseteq N$ be the set of sampled genomes.
% (but see Section~\ref{sec-retro-pro} for some subtlety around this point).
Then, $E$ is the set of edges, where each element
is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals
over which genome $c$ inherits from $p$.
Thus, each topological connection between
a parent and child node in the graph is annotated with a set of
inheritance intervals $I$.
Here, the terms parent and child are used in graph sense;
these nodes respectively represent ancestor and descendant genomes,
which can be separated by multiple generations.
We will use these two sets of terms interchangeably.

In many settings, nodes are dated, i.e.\ each
node $u\in N$ is associated with a time $\tau_u$.
The topological relationships
defined by the directed graph structure, however,
 is sufficient for many applications,
and we will assume that node dates are not known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata with nodes such as sample identifiers or quality-control metrics;
It is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.
Concretely defining nodes as integers in this way has some
significant advantages. In particular, it provides a direct link
between an ARG and its local trees in terms of a well-defined
combinatorial object (Section~\ref{sec-ARG-and-local-trees}),
as well as enabling straightforward connection between
definitions and practical implementation (Section~\ref{sec-efficiency}).

@@

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The figure shows the pedigree of eight diploid individuals, along with paths
of genetic inheritance and constituent genomes.
To avoid confusion with genomic intervals, nodes are
labelled with lowercase alphabetical letters
rather than integer identifiers.
Thus, individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the genomes
% FIXME "contained within" is a poor phrase here, use "constituent" or something
contained within a parent.
In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant product of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations (A), the corresponding gARG (B), and finally the corresponding
sequence of local trees  along the
genome (C).
% Jerome: this feels unecessary and "genealogy" is often interpreted as
% "pedigree" so needs to be used very carefully.
% The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
% which changes between local trees due to recombination.
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes.
See Section~\ref{sec-ARG-and-local-trees} for further details on
how the local trees are embedded in an ARG.
% Following the above definitions, corresponding data encoding for
% this gARG is shown in Tab.~\ref{tab-gARG-data}.

@@

\section{Event ARGs}
\label{sec-eARG}
% % I feel like sophisticated readers would at this point be saying,
% % ah, yeah, that's basically what an ARG is just in different words;
% % what's the point of this paper?
The genome ARG framework defined in the previous section is
in many ways a clarification of existing treatments
\cite[e.g.][]{mathieson2020ancestry,shipilina2023origin},
adding concrete details to describe the
differential inheritance of genetic material between genomes.
% This is definitely needed for the Nicks, who are always thinking
% about the deeper processes and are quite happy being loosey-goosey
% with representation details
Note that here, and throughout, we are not questioning the form of the
actual ancestral processes that occur in nature, but rather how we
\emph{represent} the outcomse of such processes in a practical sense.
% Again, need to remind the reader that there's an actual point
% to all this getting stuck into the details.
These practical details,
as demonstrated through the remainder of the article, have important
consequences not only for how methods exchange information about
simulated and inferred ARGs, but more fundamentally in
how we set our goals for inference and evaluate the success of results.

@@

% The prevailing view is that ARGs are composed of events, and so we
% characterise this view
% Besides viewing ARGs as being composed of events, another
% feature of the definitions given above is that they do not
% explicitly describe
% % TODO check this! Do they talk about how trees are embedded?
% the details of how such events are represented.
Where the concrete details of an ARG data structure
(the encoding) are
% TODO more?
specified~\citep[e.g.][]{
rasmussen2014genome,gusfield2014recombinatorics,ignatieva2021kwarg}
they follow the approach described by Griffiths and colleagues.
Because it is focused on recording the details of events,
we will refer to this Griffiths encoding as an ``Event ARG'', or eARG.
In an eARG, we have two different types of node in the graph
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints. These breakpoints
are used to construct the local trees.
This is done by tracing rootwards through the graph from the samples,
taking decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}.
Fig.~\ref{fig-event-arg} shows an example of a classical
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
We assume here that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

While the Griffiths approach of annotating recombination nodes
with a breakpoint in an eARG is a concise and elegant way of describing
realisations of the coalescent, it has some significant
issues when applied to the complexities of present-day datasets.
Because we are explicitly modelling the \emph{properties} of events,
the encoding must be extended to incorporate
different types of event that are not simple crossover recombination
or common ancestry.
Processes such as gene conversion, for example, can in principle
be included, but either require additional node types
or, for instance, multiple recombination nodes joined by a zero-length edge.
Modelling multiple recombinant offspring of the same parents
(while keeping track of their identity as siblings)
as occurs
in Wright-Fisher models (e.g.\ Fig.~\ref{fig-arg-in-pedigree}),
or densely sampled multi-generational
data~\citep[e.g.][]{turnbull2018hundred,hayes20191000,RosFreixedes2022}
% circumlocutions isn't technically correct, but I like the roundabout
% overly verbose feel it gives.
would require more involved circumlocutions.
These and other problems can be worked around
(although, we would ideally like everyone to use the
\emph{same} workarounds if methods are to interoperate; see the Discussion),
but it is much simpler to use the gARG encoding described
in the previous section.
By recording the \emph{outcome} of events in terms of the
intervals of genetic inheritance on edges instead of recording the
properties of the events themselves, we remove the need to explicitly
model different types of event in the data structure.
More fundamentally, because gARGs
allow us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant,
we can encode the outcome of
\emph{multiple} genetic inheritance events, and to
systematically describe the uncertainty around their
ordering (see Section~\ref{sec-precision}).

@@

Any eARG can be converted to a sample-resolved gARG using the same machinery,
and the two-step process is illustrated in Fig.~\ref{fig-ancestry-resolution}.
The first step is to take the input eARG (Fig.~\ref{fig-ancestry-resolution}A)
and convert all event nodes to genome nodes and add inheritance annotations to
the gARG's edges (Fig.~\ref{fig-ancestry-resolution}B).
If the node is a common ancestor node, we annotate the single
outbound edge with the interval $[0,L)$, for a genome of length $L$. If the
node is a recombination event with a breakpoint $x$, we annotate the two
outbound edges respectively with the intervals $[0, x)$ and $[x, L)$. These
inheritance interval annotations are clearly in one-to-one correspondence with
the information in the input eARG. They are also analogous to the
inheritance intervals we get on the edges in a prospective ARG
produced by a forwards-time simulation, which are concerned with recording
the local genetic relationship between a parent and child genome and not
necessarily minimal in terms of the resolved ancestral material of a sample.
Thus, the final step is to use the ``simplify'' algorithm to perform this
resolution~(Fig.\ref{fig-ancestry-resolution}C).
